---
title: "Class 9: Candy Mini-Project"
author: "Dea Sinaga (PID: A17725676)"
format: pdf
toc: true
---

## Background

Today we are taking a detour to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA).

## Importing candy data

The data is all realted to Halloween candy and is from the 538 website.

```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
head(candy)
```

> **Q1.** How many different candy types are in this dataset?

```{r}
nrow(candy)
```

85 different candy types are in the dataset.

> **Q2.** How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types.

> **Q3.** What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)
candy %>%
  filter(row.names(candy)=="Milky Way") |> 
  select(winpercent)
```

The winpercent value of "Milky Way" is 73.09956.

> **Q4.** What is the winpercent value for “Kit Kat”?

```{r}
candy %>%
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

The winpercent value is 76.7686.

> **Q5.** What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy %>%
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

The winpercent value is 49.6535.

There is a useful 'skim()' function in the **skimr** package that can help give you a quick overview of a given dataset. Let's install this package and try it on our candy data.

```{r}
library(skimr)
skim(candy)
```

> **Q6.** Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent seems to be on a different scale.

> **Q7.** What do you think a zero and one represent for the candy$chocolate column?

They represent whether the candy is chocolate or not. If the value is zero, then it's not chocolate. If the value is one, it is a chocolate.

## Exploratory analysis

> **Q8.** Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> **Q9.** Is the distribution of winpercent values symmetrical?

No

> **Q10.** Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
Below 50%

> **Q11.** On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.ind <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruity.ind <- as.logical(candy$fruity)
fruity.candy <- candy[fruity.ind, ]
fruity.win <- fruity.candy$winpercent
mean(fruity.win)
```

Chocolate candy is ranked higher on average than fruity candy.

> **Q12.** Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```
Yes, this difference is statistically significant.


## Overall Candy Rankings

> **Q13.** What are the five least liked candy types in this set?

```{r}
x <- c(5, 10, 1)
sort(x)
```

```{r}
order(x)
```

```{r}
candy[order(candy$winpercent), ]
```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbuster.

> **Q14.** What are the top 5 all time favorite candy types out of this set?

```{r}
candy[order(candy$winpercent, decreasing = TRUE), ]
```
Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

> **Q15.** Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> **Q16.** This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

> **Q17.** What is the worst ranked chocolate candy?

```{r, fig.height=15, fig.width=10}
mycols <- rep("gray", nrow(candy)) 

## chocolate candy in chocolate color
mycols[as.logical(candy$chocolate)] <- "chocolate"
## candy bars in brown
mycols[as.logical(candy$bar)] <- "brown"
## fruity candy in pink
mycols[as.logical(candy$fruity)] <- "pink"

ggplot(candy) +
  aes(winpercent,
      reorder(rownames(candy), winpercent)) +
        geom_col(fill=mycols)
```
Sixlets is the worst ranked chocolate candy.

> **Q18.**  What is the best ranked fruity candy?

Starburst is the best ranked fruity candy.


## Taking a look at pricepercent

Make a plot of winprcent vs pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, 
      pricepercent, 
      label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps=7)
```

> **Q19.** Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Based on the figure above, "Reese's Miniatures" is the highest ranked in terms of winpercent for the least money.

> **Q20.** What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

From the figure, I can tell that Nik L Nip, Nestle Smarties, and Ring pop are the most expensive. I can't clearly tell which ones are the next most expensive out of these: Mr Good Bar, Hershey's Milk Chocolate, Hershey's Special Dark, or Hershey's Krackel. Out of those, Nik L Nip is the least popular.


## Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns.

```{r}
cij <- cor(candy)
cij
```

```{r}
library(corrplot)
corrplot(cij)
```

> **Q22.** Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are anti-correlated.

> **Q23.** Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

I expect fruity and chocolate to have the largest contributions.


## Principal Component Analysis

In this case, we want to be sure to set 'scale=TRUE' argument for 'prcomp()' because we have one variable 'winpercent' that is on a very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

First major result figure is the "score plot" of PC1 and PC2 -- how different candy are related to each other on our new PC axis:

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(size=3.3, col=mycols, max.overlaps=7)
```

> **Q24.** Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The second major results figure from PCA.

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1)) + 
      geom_col()
```

Fruity, pluribus, and hard are picked up strongly by PC1 in the positive direction. It makes sense as in the previous figure, fruity candies are on the far right (positive direction) of PC1.

## Summary

> **Q25.** Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The characteristics that appear to make a "winning" candy are chocolate and bar, not fruity candies. The visualization figure shows that they are the highest ranked candies. The correlation plot shows that chocolate and fruity are negatively correlated with each other. The PCA plot shows that chocolate and fruity candies tend to be on opposite sides of PC1.