---
title: 'Class 12: Population Scale Analysis'
author: 'Dea Sinaga (PID: A17725676)'
date: "2026-05-11"
output: pdf_document
---

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale.

> **Q13.** Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have?

```{r}
expr <- read.table("https://bioboot.github.io/bimm143_S26/class-material/rs8067378_ENSG00000172057.6.txt")

head(expr)
```

```{r}
nrow(expr)
```

The sample size for each genotype is shown below.
```{r}
table(expr$geno)
```

```{r}
summary(expr)
```

The median expression levels for each genotypes are shown below.
```{r}
tapply(expr$exp, expr$geno, median)
```


> **Q14.** Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

The boxplot shows that the expression levels of ORMDL3 in G/G individuals are lower than that of A/A individuals. Thus, the SNP affects the expression of ORMDL3, more specifically by decreasing it.