---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Dea Sinaga (PID: A17725676)"
format: pdf
toc: true
---

## Background

Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis. 

## 1. Investigating pertussis cases by year

> **Q1.** With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

Note: The link to the cdc pertussis data no longer works but it will take you to a page on the CDC website with a graph about half-way down, under which is a Download Data (CSV) link that has the data. You can either try to use “datapasta” as suggested in the lab to extract the data into your cdc data frame or you can download the linked file and use read.csv() to upload it into a cdc data frame. 

```{r}
cdc <- read.csv("U.S. Reported Pertussis Cases_ 1922 - 2025.csv")

cdc$Number.of.Reported.Pertussis.Cases <- as.numeric(gsub("," , "" , cdc$Number.of.Reported.Pertussis.Cases))
```

```{r}
head(cdc)
```

```{r}
library(ggplot2)
```


```{r}
ggplot(cdc) +
  aes(x = Year, y = Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis Cases by Year (1922 - 2019)",
       x = "Year",
       y = "Number of Cases")
```


## 2. A tale of two vaccines (wP & aP)

Two types of pertussis vaccines have been developed: whole-cell pertussis (wP) and acellular pertussis (aP). The first vaccines were composed of ‘whole cell’ (wP) inactivated bacteria. The latter aP vaccines use purified antigens of the bacteria

> **Q2.** Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x = Year, y = Number.of.Reported.Pertussis.Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col = "blue", linetype = "dotted") +
  geom_vline(xintercept = 1996, col = "red", linetype = "dotted") +
  annotate("text", x = 1950, y = max(cdc$Number.of.Reported.Pertussis.Cases),
           label = "wP", col = "blue") +
  annotate("text", x = 2000, y = max(cdc$Number.of.Reported.Pertussis.Cases),
           label = "aP", col = "red") +
  labs(title = "Pertussis Cases by Year (1922 - 2019)",
       x = "Year",
       y = "Number of Cases")
```

I noticed from the graph above that the number of cases greatly decreased after using wP vaccine, but slightly increased again after the switch to aP.

> **Q3.** Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, the number of cases increased. One possible explanation is that people hesitated to get vaccines and that the bacteria evolved to become immune ot the vaccine.


## 3. Exploring CMI-PB data

Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis.

**The CMI-PB API returns JSON data**

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
```

```{r}
head(subject, 3)
```

> **Q4.** How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 87 aP and 85 wP infant vaccinated subjects.

> **Q5.** How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 112 female patients and 60 male patients.

> **Q6.** What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
tab <- table(subject$biological_sex, subject$race)
knitr::kable(tab)
```

**Side-Note: Working with dates**

> **Q7.** Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)
```

(i) average age of wP individuals
```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)
```

```{r}
wP <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length(wP$age, "years")))
```

The average age of wP individuals is 37.

(ii) average age of aP individuals

```{r}
aP <- subject %>% filter(infancy_vac == "aP")
round(summary(time_length(aP$age, "years")))
```

The average age of aP individuals is 28.

(iii) are they significantly different

```{r}
t.test(wP$age, aP$age)
```

They are significantly different.

> **Q8.** Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> **Q9.** With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

Yes, they are significantly different.

**Joining multiple tables**

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 

titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```

> **Q9.** Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- left_join(specimen, subject)
dim(meta)
head(meta)
```

> **Q10.** Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
head(abdata)
```

> **Q11.** How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> **Q12.** What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The most "recent" dataset has the least number of rows.


## 4. Examine IgG Ab titer levels

Now using our joined/merged/linked abdata dataset filter() for IgG isotype.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> **Q13.** Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> **Q14.** What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

Based on the box plot above, PT, PRN, FIM2/3, and FHA show clear differences in the level of IgG antibody titers. The booster contains pertussis-related antigens, so the antigen levels targeted by the vaccines would increase over time.

> **Q15.** Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "OVA antigen levels per visit (ap red, wP teal)")
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw() +
  labs(title = "PT antigen levels per visit (ap red, wP teal)")
```

> **Q16.** What do you notice about these two antigens time courses and the PT data in particular?

PT levels increase over time and exceeds OVA. It looks like PT levels (both for aP and wP) peaked at visit 5 and slightly decreased after.

> **Q17.** Do you see any clear difference in aP vs. wP responses?

No, the responses for aP and wP seem to have a similar trend.

Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> **Q18.** Does this trend look similar for the 2020 dataset?

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

No, it doesn't look similar.


## 5. Obtaining CMI-PB RNASeq data

For RNA-Seq data the API query mechanism quickly hits the web browser interface limit for file size. However, we can still do “targeted” RNA-Seq querys via the web accessible API.

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> **Q19.** Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> **Q20.** What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

It's at its maximum level around the 4th and 8th visits.

> **Q21.** Does this pattern in time match the trend of antibody titer data? If not, why not?

Not exactly because antibodies are generated later than the ssRNA response.
